Gaze-dependent evidence accumulation predicts multi-alternative risky choice behaviour

### 2021-02-05 manuscript/main.tex
revised supplement sections

\section*{Acknowledgements}
F.M. was supported by the International Max Planck Research School on the Life Course (LIFE). A.T. is supported by Stanford Data Science.

@@

\subsection*{Supplementary Note 1: Additional eye movement analyses}
\label{sup:eye-movement-analyses}

@@

% Supplementary table: Summary of GLA estimates
\subsection*{GLA parameter estimates}
\begin{table}[h!]
\begin{centering}
\begin{tabular}{@{}llllllll@{}}
\toprule
       & mean & SD   & min  & 25\% & 50\% & 75\% & max   \\ \midrule
$\alpha$  & 0.47 & 0.35 & 0.05 & 0.25 & 0.37 & 0.61 & 1.67  \\
$\beta$   & 6.81 & 9.87 & 0.04 & 1.13 & 3.25 & 8.04 & 49.74 \\
$\gamma$  & 0.81 & 0.25 & 0.22 & 0.58 & 1.0  & 1.0  & 1.0   \\
$\lambda$ & 0.29 & 0.2  & 0.08 & 0.14 & 0.23 & 0.46 & 0.65  \\
$\theta$  & 0.69 & 0.18 & 0.13 & 0.63 & 0.72 & 0.83 & 0.95  \\ \bottomrule
\end{tabular}
\caption{\textbf{Summary of GLA estimates.} $\alpha$ is the utility parameter. $\beta$ is the inverse temperature parameter of the choice rule (0 = random choice). $\gamma$ is the probability weighting parameter (1 = objective probability weighting). $\lambda$ is the leak parameter (0 = perfect memory, 1 = full leak of all previous information). $\theta$ is the gaze-discount parameter (1 = no gaze-discount, 0 = maximum gaze-discount).}
\label{tab:gla-estimates}
\end{centering}
\end{table}
\clearpage

@@

\subsection*{Context-effect strength and individual relative model fit}

% Supplementary Fig.: Relative model fit and RST
\begin{figure}[ht!]
\begin{centering}
\includegraphics[scale=1]{../figures/S_GLA-vs-MDFT_dBIC_RST.pdf}
\caption{\textbf{Relative model fits between GLA and MDFT in relation to RST}. Relative model fits of MDFT (indicated by BIC difference between GLA and MDFT) tended to be higher for participants with higher RST in attraction trials (left panel; slope = 0.04, HDI$_{95}$ = [-0.01, 0.09] increase in RST per 100 unit increase in BIC difference, 93.6\% of posterior mass above 0), but not compromise trials (right panel), even though 7 out of 9 participants with attraction RST above 0.7 were better described by GLA overall (participants left of dashed vertical line).}
\label{fig:dbic-rst}
\end{centering}
\end{figure}
\newpage

@@

\subsection*{Best average fitting switchboard variants}
% Supplementary Table: Best average fitting (mean BIC) switchboard variants
\begin{table}[!ht]
\begin{centering}
\begin{tabular}{@{}cccccccc@{}}
\toprule
\textbf{Rank} & $\mathbf{GD_{Alt}}$                 & $\mathbf{GD_{Att}}$                 & \textbf{Leak}                         & \textbf{Inhibition}                   & \textbf{Integration}    & \textbf{Comparison}    & \textbf{BIC}    \\ \midrule
1    & \cellcolor[HTML]{DAE8FC}\textbf{Yes} & No                          & Constant                     & None                         & Multiplicative & \textit{n.d.} & 232.08 \\
2    & No                          & No                          & Constant                     & \cellcolor[HTML]{DAE8FC}\textbf{Gaze} & Multiplicative & Independent   & 235.94 \\
3    & \cellcolor[HTML]{DAE8FC}\textbf{Yes} & No                          & Constant                     & \cellcolor[HTML]{DAE8FC}\textbf{Gaze}                         & Multiplicative & Comparative   & 236.75 \\
4    & \cellcolor[HTML]{DAE8FC}\textbf{Yes} & \cellcolor[HTML]{DAE8FC}\textbf{Yes} & Constant                     & None                         & Multiplicative & \textit{n.d.} & 237.21 \\
5    & \cellcolor[HTML]{DAE8FC}\textbf{Yes} & No                          & Constant                     & \cellcolor[HTML]{DAE8FC}\textbf{Gaze} & Multiplicative & Independent   & 237.31 \\
6    & \cellcolor[HTML]{DAE8FC}\textbf{Yes} & No                          & Constant                     & Constant                     & Multiplicative & \textit{n.d.} & 237.40 \\
7    & \cellcolor[HTML]{DAE8FC}\textbf{Yes} & No                          & \cellcolor[HTML]{DAE8FC}\textbf{Gaze} & None                         & Multiplicative & Comparative   & 238.41 \\
8    & No                          & \cellcolor[HTML]{DAE8FC}\textbf{Yes} & Constant                     & \cellcolor[HTML]{DAE8FC}\textbf{Gaze} & Multiplicative & Independent   & 241.04 \\
9    & \cellcolor[HTML]{DAE8FC}\textbf{Yes} & No                          & Constant                     & Distance                     & Multiplicative & Comparative   & 241.40 \\
10   & \cellcolor[HTML]{DAE8FC}\textbf{Yes} & \cellcolor[HTML]{DAE8FC}\textbf{Yes} & Constant                     & \cellcolor[HTML]{DAE8FC}\textbf{Gaze} & Multiplicative & Comparative   & 241.78 \\ \bottomrule
\end{tabular}
\caption{\textbf{Overview of average best fitting model variants.} All ten model variants that fit the data best on average used some form of gaze-dependence (blue shaded cells), mostly an alternative-wise gaze discount. \textit{"n.d."} denotes variants where comparison mechanisms were not distinguishable by the analysis.}
\label{tab:switchboard-best-agg}
\end{centering}
\end{table}
\clearpage

\subsection*{Counts of individual best fitting switch levels}
% Supplementary Fig.: Individual best fitting switch level counts
\begin{figure}[!ht]
\begin{centering}
\includegraphics[scale=1]{../figures/S_switch-level_individual_counts.pdf}
\caption{\textbf{Counts of individual best fitting switches.} Most participants were best described by model variants that included multiplicative attribute integration, with alternative-wise gaze discount, no attribute-wise gaze discount, constant leakage and no inhibition.}
\label{fig:switchboard-ind-switch-counts}
\end{centering}
\end{figure}
\newpage

\subsection*{GLA and hybrid variant predicted choice probabilities by dwell time advantage for weak and strong attraction responders}
% Supplementary Fig.: Relative dwell advantage for GLA and hybrid, subgroups (split by attraction effect strength)
\begin{figure}[!ht]
\begin{centering}
\includegraphics[scale=1]{../figures/S_dwell-advantage_subgroups_gla-hybrid.pdf}
\caption{\textbf{Observed and model-predicted association of dwell time advantage and choice for participants with weaker and strong attraction effects.} \textbf{(a-f)} Data and model predictions for participants with weaker attraction effects (RST $<$ 0.7). \textbf{(g-l)} Data and model predictions for participants with strong attraction effects (RST $>$ 0.7) Each column refers to one choice alternative: Target (first column; \textbf{a, d, g, j}); Competitor (second column; \textbf{b, e, h, k}); Decoy (third column; \textbf{c, f, i, l}). Rows refer to trials in attraction \textbf{(a-c, g-i)} and compromise trials \textbf{(d-f, j-l)}. White and grey bars and error bars show observed mean $\pm$ s.e. choice probabilities computed from even-numbered trials, for participants with weaker and stronger attraction effects, respectively. Coloured lines indicate model predictions derived from 50 simulations for each odd-numbered trial.}
\label{fig:rdt-adv-subgroups}
\end{centering}
\end{figure}
\newpage

\subsection*{Supplementary Note 2: No process evidence that strong attraction responders follow simple choice rule}
\label{sup:ade-choicerule}

Using process measures, we performed multiple tests of the hypothesis, that individuals with strong attraction effects follow a simple choice rule of choosing the dominant alternative. First, we tested whether the strength of individual attraction effects (individual RST in attraction trials) was related to differences in mean response times (RTs) in attraction trials. If individuals used a choice rule, their choices might be made faster, as they do not engage in multiple pairwise comparisons or calculations of expected outcomes. There was no correlation between the two measures ($r$ = 0.06, HDI$_{95}$ = [-0.24, 0.34]). Similarly, no relationship was found between individual RST and the number of fixations in attraction trials ($r$ = 0.06, HDI$_{95}$ = [-0.25, 0.35]). Mean RTs in attraction trials did not meaningfully differ between trials with target choices and trials with other choices ($d$ = -0.2, HDI$_{95}$ = [-0.67, 0.25]). Next, we tested whether individuals with strong attraction effects committed to a choice once they learned about the dominance relationship in the stimuli, as if using the dominance relationship as a stopping rule, or if they kept exploring the stimuli. There was, however, no relationship between individual RST and the mean number of fixations after all target and decoy attributes were fixated at least once ($r$ = 0.04, HDI$_{95}$ = [-0.28, 0.31]). The same analysis using fixation counts after target and decoy alternatives were both seen at least once on any attribute revealed no effect either ($r$ = 0.11, HDI$_{95}$ = [-0.20, 0.40]).
Taken together, we did not find any evidence based on process data to support the hypothesis that strong attraction responders used a simple choice rule.

% Supplement references
\renewcommand*{\bibfont}{\small}
\printbibliography[title=Supplementary References]

>>>>

\section*{Acknowledgements}
F.M. is supported by the International Max Planck Research School on the Life Course (LIFE). A.T. is supported by Stanford Data Science.

@@

\subsection*{Supplementary Note 1}
\label{sup:eye-movement-analyses}

@@

% Supplementary table: Summary of GLA estimates
\begin{table}[h!]
\begin{centering}
\begin{tabular}{@{}llllllll@{}}
\toprule
       & mean & SD   & min  & 25\% & 50\% & 75\% & max   \\ \midrule
$\alpha$  & 0.47 & 0.35 & 0.05 & 0.25 & 0.37 & 0.61 & 1.67  \\
$\beta$   & 6.81 & 9.87 & 0.04 & 1.13 & 3.25 & 8.04 & 49.74 \\
$\gamma$  & 0.81 & 0.25 & 0.22 & 0.58 & 1.0  & 1.0  & 1.0   \\
$\lambda$ & 0.29 & 0.2  & 0.08 & 0.14 & 0.23 & 0.46 & 0.65  \\
$\theta$  & 0.69 & 0.18 & 0.13 & 0.63 & 0.72 & 0.83 & 0.95  \\ \bottomrule
\end{tabular}
\caption{\textbf{Summary of GLA estimates.} $\alpha$ is the utility parameter. $\beta$ is the inverse temperature parameter of the choice rule (0 = random choice). $\gamma$ is the probability weighting parameter (1 = objective probability weighting). $\lambda$ is the leak parameter (0 = perfect memory, 1 = full leak of all previous information). $\theta$ is the gaze-discount parameter (1 = no gaze-discount, 0 = maximum gaze-discount).}
\label{tab:gla-estimates}
\end{centering}
\end{table}
\clearpage

@@

% Supplementary Fig.: Relative model fit and RST
\begin{figure}[ht!]
\begin{centering}
\includegraphics[scale=1]{../figures/S_GLA-vs-MDFT_dBIC_RST.pdf}
\caption{\textbf{Relative model fits between GLA and MDFT in relation to RST}. Relative model fits of MDFT (indicated by BIC difference between GLA and MDFT) tended to be higher for participants with higher RST in attraction trials (left panel; slope = 0.04, HDI$_{95}$ = [-0.01, 0.09] increase in RST per 100 unit increase in BIC difference, 93.6\% of posterior mass above 0), but not compromise trials (right panel), even though 7 out of 9 participants with attraction RST above 0.7 were better described by GLA overall (participants left of dashed vertical line).}
\label{fig:dbic-rst}
\end{centering}
\end{figure}
\clearpage

@@

% Supplementary Table: Best average fitting (mean BIC) switchboard variants
\begin{table}[!ht]
\begin{centering}
\begin{tabular}{@{}cccccccc@{}}
\toprule
\textbf{Rank} & $\mathbf{GD_{Alt}}$                 & $\mathbf{GD_{Att}}$                 & \textbf{Leak}                         & \textbf{Inhibition}                   & \textbf{Integration}    & \textbf{Comparison}    & \textbf{BIC}    \\ \midrule
1    & \cellcolor[HTML]{DAE8FC}\textbf{Yes} & No                          & Constant                     & None                         & Multiplicative & \textit{n.d.} & 232.08 \\
2    & No                          & No                          & Constant                     & \cellcolor[HTML]{DAE8FC}\textbf{Gaze} & Multiplicative & Independent   & 235.94 \\
3    & \cellcolor[HTML]{DAE8FC}\textbf{Yes} & No                          & Constant                     & \cellcolor[HTML]{DAE8FC}\textbf{Gaze}                         & Multiplicative & Comparative   & 236.75 \\
4    & \cellcolor[HTML]{DAE8FC}\textbf{Yes} & \cellcolor[HTML]{DAE8FC}\textbf{Yes} & Constant                     & None                         & Multiplicative & \textit{n.d.} & 237.21 \\
5    & \cellcolor[HTML]{DAE8FC}\textbf{Yes} & No                          & Constant                     & \cellcolor[HTML]{DAE8FC}\textbf{Gaze} & Multiplicative & Independent   & 237.31 \\
6    & \cellcolor[HTML]{DAE8FC}\textbf{Yes} & No                          & Constant                     & Constant                     & Multiplicative & \textit{n.d.} & 237.40 \\
7    & \cellcolor[HTML]{DAE8FC}\textbf{Yes} & No                          & \cellcolor[HTML]{DAE8FC}\textbf{Gaze} & None                         & Multiplicative & Comparative   & 238.41 \\
8    & No                          & \cellcolor[HTML]{DAE8FC}\textbf{Yes} & Constant                     & \cellcolor[HTML]{DAE8FC}\textbf{Gaze} & Multiplicative & Independent   & 241.04 \\
9    & \cellcolor[HTML]{DAE8FC}\textbf{Yes} & No                          & Constant                     & Distance                     & Multiplicative & Comparative   & 241.40 \\
10   & \cellcolor[HTML]{DAE8FC}\textbf{Yes} & \cellcolor[HTML]{DAE8FC}\textbf{Yes} & Constant                     & \cellcolor[HTML]{DAE8FC}\textbf{Gaze} & Multiplicative & Comparative   & 241.78 \\ \bottomrule
\end{tabular}
\caption{\textbf{Overview of average best fitting model variants.} All ten model variants that fit the data best on average used some form of gaze-dependence (blue shaded cells), mostly an alternative-wise gaze discount. \textit{"n.d."} denotes variants where comparison mechanisms were not distinguishable by the analysis.}
\label{tab:switchboard-best-agg}
\end{centering}
\end{table}
\clearpage

% Supplementary Fig.: Individual best fitting switch level counts
\begin{figure}[!ht]
\begin{centering}
\includegraphics[scale=1]{../figures/S_switch-level_individual_counts.pdf}
\caption{\textbf{Counts of individual best fitting switches.} Most participants were best described by model variants that included multiplicative attribute integration, with alternative-wise gaze discount, no attribute-wise gaze discount, constant leakage and no inhibition.}
\label{fig:switchboard-ind-switch-counts}
\end{centering}
\end{figure}
\clearpage

% Supplementary Fig.: Relative dwell advantage for GLA and hybrid, subgroups (split by attraction effect strength)
\begin{figure}[!ht]
\begin{centering}
\includegraphics[scale=1]{../figures/S_dwell-advantage_subgroups_gla-hybrid.pdf}
\caption{\textbf{Observed and model-predicted association of dwell time advantage and choice for participants with weaker and strong attraction effects.} \textbf{(a-f)} Data and model predictions for participants with weaker attraction effects (RST $<$ 0.7). \textbf{(g-l)} Data and model predictions for participants with strong attraction effects (RST $>$ 0.7) Each column refers to one choice alternative: Target (first column; \textbf{a, d, g, j}); Competitor (second column; \textbf{b, e, h, k}); Decoy (third column; \textbf{c, f, i, l}). Rows refer to trials in attraction \textbf{(a-c, g-i)} and compromise trials \textbf{(d-f, j-l)}. White and grey bars and error bars show observed mean $\pm$ s.e. choice probabilities computed from even-numbered trials, for participants with weaker and stronger attraction effects, respectively. Coloured lines indicate model predictions derived from 50 simulations for each odd-numbered trial.}
\label{fig:rdt-adv-subgroups}
\end{centering}
\end{figure}
\clearpage

\subsection*{Supplementary Note 2}
\label{sup:ade-choicerule}
\subsubsection*{No process evidence that strong attraction responders follow simple choice rule}
Using process measures, we performed multiple tests of the hypothesis, that individuals with strong attraction effects follow a simple choice rule of choosing the dominant alternative. First, we tested whether the strength of individual attraction effects (individual RST in attraction trials) was related to differences in mean response times (RTs) in attraction trials. If individuals used a choice rule, their choices might be made faster, as they do not engage in multiple pairwise comparisons or calculations of expected outcomes. There was no correlation between the two measures ($r$ = 0.06, HDI$_{95}$ = [-0.24, 0.34]). Similarly, no relationship was found between individual RST and the number of fixations in attraction trials ($r$ = 0.06, HDI$_{95}$ = [-0.25, 0.35]). Mean RTs in attraction trials did not meaningfully differ between trials with target choices and trials with other choices ($d$ = -0.2, HDI$_{95}$ = [-0.67, 0.25]). Next, we tested whether individuals with strong attraction effects committed to a choice once they learned about the dominance relationship in the stimuli, as if using the dominance relationship as a stopping rule, or if they kept exploring the stimuli. There was, however, no relationship between individual RST and the mean number of fixations after all target and decoy attributes were fixated at least once ($r$ = 0.04, HDI$_{95}$ = [-0.28, 0.31]). The same analysis using fixation counts after target and decoy alternatives were both seen at least once on any attribute revealed no effect either ($r$ = 0.11, HDI$_{95}$ = [-0.20, 0.40]).
Taken together, we did not find any evidence based on process data to support the hypothesis that strong attraction responders used a simple choice rule.

% Supplement references
\renewcommand*{\bibfont}{\small}
\printbibliography[title=Supplementary Information References]
